Analysis-ready VCF at Biobank scale using Zarr

### 2025-01-25 paper.tex

\begin{table}
\caption{Summary for VCF Zarr conversion of 
All of Us exome-like data genotype data for chromosome 22 (245,394 samples, 
715,256 variants) 
consisting of 25 fields and 8.5GiB of storage ($\sim1.1$X larger
% 8.5 / 7.44 = 1.142
than source gzipped VCF).
Here we use the local alleles fields \texttt{call\_LA} and \texttt{call\_LAD};
see text for details.
For column details, see Table~\ref{tab-genomics-england-data}.
\label{tab-auo-data}}
\begin{tabular}{llS[table-format=3.2]S[table-format=4.1]S[table-format=3.2]}
\toprule
{Field} & {Type} & {Storage} & {Compress} & {\% Total} \\
\midrule
/call\_LAD & int16 & 2.26 GiB & 290.0 & 27.85\% \\
/call\_GQ & int8 & 1.89 GiB & 87.0 & 23.29\% \\
/call\_LA & int8 & 1.24 GiB & 260.0 & 15.28\% \\
/call\_genotype & int8 & 914.22 MiB & 370.0 & 11.00\% \\
/call\_RGQ & int16 & 729.57 MiB & 460.0 & 8.78\% \\
/call\_genotype\_mask & bool & 606.02 MiB & 550.0 & 7.29\% \\
/call\_FT & object & 498.61 MiB & 2700.0 & 6.00\% \\
\bottomrule
\end{tabular}
\end{table}

@@

@@

To demonstrate the versatility of \texttt{vcf2zarr}, in this section
we include a case study of a dataset originating from a species whose
genome properties differ substantially from human. The conifer Norway
spruce, \emph{Picea abies}, is one of the largest and most
ecologically important species on Earth and is distributed over large
parts of the Northern hemisphere. One of its features is its large
genome, consisting of 12 autosomes with a genome size of 19.6
Gb~\cite{nystedt_NorwaySpruceGenome_2013}. Here, we assess the
performance of \texttt{vcf2zarr} as applied to 1,063 whole-genome
resequenced spruce individuals from a recent study (in preparation),
where resequencing data has been mapped to a chromosome-scale
reference genome. The dataset consists of approximately 3,745 million
single-nucleotide variants and small indels from the 12 autosomes
totalling 7.4TiB of VCF data after \texttt{bgzip} compression,
distributed over 165 VCF files. Since the dataset originally was
generated for downstream analyses that require genotype likelihoods,
the genotypes include the PL field.

@@

\begin{table}
\caption{Summary for a selection of the largest VCF Zarr columns
produced for 1,063 VCFs on 12 spruce chromosomes using
\texttt{vcf2zarr} with samples chunk size set to 1,063, variants
chunk size to 10,000, and using local allele fields to reduce output
size. For column details see the caption of
Table~\ref{tab-genomics-england-data}.
\label{tab-spruce-data}}

>>>>

\begin{table}
\caption{Summary for VCF Zarr conversion of 
All of Us exome-like data genotype data for chromosome 22 (245,394 samples, 
715,256 variants) 
consisting of 25 fields and 8.5GiB of storage ($\sim1.1$X larger
% 8.5 / 7.44 = 1.142
than source gzipped VCF; see text for discussion).
Here we use the local alleles fields \texttt{call\_LA} and \texttt{call\_LAD}; 
see text for details.
For column details, see Table~\ref{tab-genomics-england-data}.
\label{tab-auo-data}}
\begin{tabular}{llS[table-format=3.2]S[table-format=4.1]S[table-format=3.2]}
\toprule
{Field} & {Type} & {Storage} & {Compress} & {\% Total} \\
\midrule
/call\_LAD & int16 & 2.26 GiB & 290.0 & 27.85\% \\
/call\_GQ & int8 & 1.89 GiB & 87.0 & 23.29\% \\
/call\_LA & int8 & 1.24 GiB & 260.0 & 15.28\% \\
/call\_genotype & int8 & 914.22 MiB & 370.0 & 11.00\% \\
/call\_RGQ & int16 & 729.57 MiB & 460.0 & 8.78\% \\
/call\_genotype\_mask & bool & 606.02 MiB & 550.0 & 7.29\% \\
/call\_FT & object & 498.61 MiB & 2700.0 & 6.00\% \\
\bottomrule
\end{tabular}
\end{table}

@@

[Discuss why we're bigger than VCF here, people will want to understand.

@@

To demonstrate the versatility of \texttt{vcf2zarr}, in this section
we include a case study of a dataset originating from a species whose
genome properties differ substantially from human. The conifer Norway
spruce, \emph{Picea abies}, is one of the largest and most
ecologically important species on Earth and is distributed over large
parts of the Northern hemisphere. One of its features is its large
genome, consisting of 12 autosomes with a genome size of 19.6
Gb~\cite{nystedt_NorwaySpruceGenome_2013}. Here, we assess the
performance of \texttt{vcf2zarr} as applied to 1,063 whole-genome
resequenced spruce individuals from a recent study (in preparation),
where resequencing data has been mapped to a chromosome-scale
reference genome. The dataset consists of approximately 3.8 million
single-nucleotide variants and small indels from the 12 autosomes
totalling 7.4TiB of VCF data after \texttt{bgzip} compression,
distributed over 165 VCF files. Since the dataset originally was
generated for downstream analyses that require genotype likelihoods,
the genotypes include the PL field.

@@

\begin{table}
\caption{Summary for VCF Zarr conversion of the Norway spruce
WGS data on all 12 autosomes (1,063 samples, 374,517,0452 variants)
consisting of 34 fields and 2.3TiB of storage ($\sim 3.2X$ smaller 
than source gzipped VCFs). The chunk size is 10,000 variants times 
1,063 samples (i..e, a single sample chunk). This uses local 
alleles to reduce the 
space required by the PL field; see the text for discussion.
For column details see the caption of
Table~\ref{tab-genomics-england-data}.
\label{tab-spruce-data}}
